---
title: "Section 7. Matched Pair Tests & Comparing Data to Standards"
output: html_notebook:
  toc: true
---

```{r}
# libraries-----------------------------------------------------------
source("NADA2_online_3_9/Class Data/Loadlibs.R")
library(tidyverse)
library(glue)
```

```{r}
# data --------------------------------------------------------
data_folder <- "NADA2_online_3_9/Class Data"

load(glue("{data_folder}/SedHg.RData"))
load(glue("{data_folder}/EppSedHg.Rda"))

# these for the exercises
load(glue("{data_folder}/Example1.RData"))
load(glue("{data_folder}/Example2.RData"))
load(glue("{data_folder}/Example3.RData"))
data(Atra)
```

# Matched pair tests

## Sign test

Whether the first is different from the second. Want the Fong correction for ties.

```{r}
as_tibble(EppSedHg)

# two sided
cen_signtest(EppSedHg$Hg2001, EppSedHg$Cens01, EppSedHg$Hg1996, EppSedHg$Cens96)

# one-sided higher
cen_signtest(EppSedHg$Hg2001, EppSedHg$Cens01, EppSedHg$Hg1996, EppSedHg$Cens96, alternative = "greater")
```

## Signed rank test

* Generally has more power than the sign test when the magnitude of difference is meaningful (most of the time).
* Computes the differences and ranks them according to absolute value.
* Ties discarded (Pratt correction for ties)
* Test statistic = sum of the positive signed ranks / standard deviation. 
* Small p.value so reject the null hypothesis and would say it is different (would expect 2001 to be higher than 1996)

```{r}
# two-sided
cen_signedranktest(EppSedHg$Hg2001, EppSedHg$Cens01, EppSedHg$Hg1996, EppSedHg$Cens96)

# one-sided higher
cen_signedranktest(EppSedHg$Hg2001, EppSedHg$Cens01, EppSedHg$Hg1996, EppSedHg$Cens96, alternative = "greater")
```

## Paired Prentice Wilcoxon test

* Non-parametric test
* more power than the sign and signed-rank tests when differences are asymmetric (often for enviro data)
* built to handle ties
* similar to a paired t-test on scores



```{r}
# two sided
ppw.test(EppSedHg$Hg2001, EppSedHg$Cens01, EppSedHg$Hg1996, EppSedHg$Cens96) # cannot reject HA. med diff is 0.015

# one.sided higher
ppw.test(EppSedHg$Hg2001, EppSedHg$Cens01, EppSedHg$Hg1996, EppSedHg$Cens96, alternative = "greater") # again cannot reject
```

## Censored paired test

* a parametric test
* a censored version of a paired t-test of means
* assumes that the paired differences have a normal dist (check with a qq plot)
* if non-normal (differences) p-values may be too high
* if p-values are low then probably fine
* one-sided LCL against zero, but could also compare to a non-zero such as a standard
* done using MLE

Output includes graphs:
* data distribution
* qq plot
* pp plot

In this example the data does not look like it is from a normal distribution

```{r}
# two sided
cen_paired(EppSedHg$Hg2001, EppSedHg$Cens01, EppSedHg$Hg1996, EppSedHg$Cens96)

# one sided also available
```

## Fixing non-normality

* use a non-parametric test
* could take logs, but then no longer testing on the mean
* permutation test on mean difference


## Comparing to a standard

* instead of a second column, just put in a number
* LCL95 compared to the standard

```{r}
# does the mean of 2001 Hg concentrations exceed a standard of 0.01?
cen_paired(EppSedHg$Hg2001, EppSedHg$Cens01, 0.1, alt = "greater")

# in this case, p-value is small so reject H0 of equality. Mean exceeds standard.
```

# Comparing data with non-detects to a standard

* Can fit a skewed distribution (MLE or rROS) to the data and test whether the data fit the standard or not
* one sided LCL95 or UCL95
* could bootstrap if > 20 observations
* if n < 20 then best fit (MLE)
* if n >= 20 then use Kaplan-Meier (non-parametric) or best fitting dist 
* if n > 70 could use t-test, but no benefit compared to bootstrap
* dist is hardly ever normal when there are non-detects
* BIC better than PPCC or Shapiro Wilk as it evaluates the entire distribution shape, the cdf.
* if negative numbers are part of fitted dist, use second highest PPCC or SW, then use the second highest instead of normal

## one sample tests

* just confidence intervals
* H0 assumes compliance when using LCL
* H0 assumes non-compliance when using UCL


## Exercises

1a. Compare Data to a Standard Using a Matched Pair Test

```{r}
# see if arsenic concentrations exceed the drinking water standard of 10 ug/L
cen_paired(Example1$Arsenic, Example1$NDisTRUE, 10, alt = "greater")
# p value is large so cannot reject H0 that Arsenic does not exceed standard
```

1b. Test for Differences in Two Paired Columns of Data

```{r}
# test whether atrazine concentrations were the same in June v September. 
# Test both for differences in the mean and well differences in the cdfs and medians

cen_paired(Atra$June, Atra$JuneCen, Atra$Sept, Atra$SeptCen) # test for mean difference
# cannot reject H0 that mean diff is zero

cen_signtest(Atra$June, Atra$JuneCen, Atra$Sept, Atra$SeptCen) # test for median
# also cannot reject H0 that median diff is zero using Fong correction

cen_signedranktest(Atra$June, Atra$JuneCen, Atra$Sept, Atra$SeptCen) # cdf test
# this has more poser to see differences.
# p-value low so can reject H0 of no difference. This test finds a diff between datasets (higher in Sep)
```

1c. Comparing Data to Standards Using an Upper Confidence Limit

```{r}
# summary
censummary(Example1$Arsenic, Example1$NDisTRUE) # 21 samples - a bit small

# boxplot
cboxplot(Example1$Arsenic, Example1$NDisTRUE, show = TRUE) # showing values below detection limit (most of them)

# compare cdfs to decide on the distribution
cenCompareCdfs(Example1$Arsenic, Example1$NDisTRUE, Yname = "Arsenic Concentration in ug/L") # gamma has lowest BIC
```

```{r}
# compute UCL95 using MLE
egam_mle <- egammaAltCensored(Example1$Arsenic,
                          Example1$NDisTRUE,
                          ci = TRUE,
                          ci.type = "upper",
                          ci.method = "normal.approx")
print(egam_mle)
```

```{r}
# compute UCL95 using bootstrap
# non parametric method chosen with bootstrap
# computes kaplan meier as we have 2/3 of data as non detects
egam_bs <- enparCensored(Example1$Arsenic, 
                          Example1$NDisTRUE,
                          ci = TRUE,
                          ci.type = "upper",
                          ci.method = "bootstrap",
                          n.bootstraps = 5000)
print(egam_bs)

```

2a. Computation of a UCL95 for data with both detected and non-detected values, DL unknown.

```{r}
# lowest reported was 0.1229 so assumed that all ND are lower than this


# MIBK is methyl isobutyl ketone
mibk_ucl95 <- elnormAltCensored(Example2$MIBK,
                                Example2$MIBKcen,
                                method = "rROS",
                                ci = TRUE,
                                ci.method = "bootstrap",
                                ci.type = "upper",
                                n.bootstraps = 5000)

print(mibk_ucl95) # not as good as with lab provided limit, but still useful to use lowest detected when estimating
```

```{r}
# with reporting limit provided by lab of 0.029
mibk2_ucl95 <- elnormAltCensored(Example2$MIBK2,
                                Example2$MIBK2cen,
                                method = "rROS",
                                ci = TRUE,
                                ci.method = "bootstrap",
                                ci.type = "upper",
                                n.bootstraps = 5000)

print(mibk2_ucl95) # better with lab provided limit
```

3a. Compute when all are ND

```{r}
# know that all detection limits are below 10 ppb for arsenic in drinking water
# compute Maximum Concentration Limit (MCL)
# 0 of 14 exceed the MCL as the detection limits are below the MCL

binom.test(0, 14, alternative = "less") # 0 successes from 14 trials in one-sided binom test
# conf int is 0 to 0.192 
# so we can say with 95% probability that no more than 19.2% of concentrations exceed the MCL
```

