---
title: "Section 11. Trend Analysis with Nondetects"
output: html_notebook:
  toc: true
---

```{r}
# libraries-----------------------------------------------------------
source("NADA2_online_3_9/Class Data/Loadlibs.R")
library(tidyverse)
library(glue)
```

```{r}
# data --------------------------------------------------------
data_folder <- "NADA2_online_3_9/Class Data"

load(glue("{data_folder}/DairyCreekChromium.RData"))
```


# Parametric Trend Analysis

* check for multicollinearity between independent variables - vif test
* check residuals to see that they follow the assumed distribution
* choose the model with the lowest AIC
* uses decimal time

## Dataset

```{r}
Dairy_Creek_Chromium

censummary(Dairy_Creek_Chromium$`Total Recoverable Chromium`, as.logical(Dairy_Creek_Chromium$CrND))
```

## Single variable

```{r}
cencorreg(Dairy_Creek_Chromium$`Total Recoverable Chromium`, Dairy_Creek_Chromium$CrND, Dairy_Creek_Chromium$dectime)
# default uses logs
# slope is significant, showing a decrease of 5.9 per cent per year
# plot and Shapiro-Francia W look fine
```

## Multiple regression, x and time

```{r}
xvar2 <- Dairy_Creek_Chromium %>% 
  dplyr::select(dectime, mean_daily_flow_cfs)

reg_cr <- cencorreg(Dairy_Creek_Chromium$`Total Recoverable Chromium`, Dairy_Creek_Chromium$CrND, xvar2) # smaller AIC

summary(reg_cr) # trend and flow significant
# decrease of 5.1 per cent per year


```

```{r}
# check vif
cr_lm <- lm(`Total Recoverable Chromium` ~ dectime + mean_daily_flow_cfs, data = Dairy_Creek_Chromium)
vif(cr_lm) # all good
```

## Seasonal regression with sine and cosine

* sine and cosine of 2 * pi * time in decimal years
* one revolution every year (2 * pi * time)
* can compare the AIC for models with and without the sine and cosine terms

```{r}
cos_t <- cos(2*pi*Dairy_Creek_Chromium$dectime)
sin_t <- sin(2*pi*Dairy_Creek_Chromium$dectime)

xvar4 <- tibble(Dairy_Creek_Chromium$dectime, Dairy_Creek_Chromium$mean_daily_flow_cfs, sin_t, cos_t)

reg4 <- cencorreg(Dairy_Creek_Chromium$`Total Recoverable Chromium`, Dairy_Creek_Chromium$CrND, xvar4) 
# AIC not as good as 2 var model
# no significant seasonal variation
# Shapiro-Francia not bad except for one point

summary(reg4)
```

```{r}
# check vif
cr_lm2 <- lm(Dairy_Creek_Chromium$`Total Recoverable Chromium` ~ 
               Dairy_Creek_Chromium$dectime + 
               Dairy_Creek_Chromium$mean_daily_flow_cfs +
               sin_t +
               cos_t)
vif(cr_lm2) # all good
```

## Nonparametric trend tests with censored data

* based on ATS line
* slope produces Kendall's tau of 0 for the residuals from the line

## Single variable non parametric with time

```{r}
# if linear in original units, then log = false otherwise log = true (default) because fitting line
ATS(Dairy_Creek_Chromium$`Total Recoverable Chromium`, Dairy_Creek_Chromium$CrND, Dairy_Creek_Chromium$dectime, LOG = FALSE) 
# won't matter for tests but good for line
# normal units may be easier to interpret depending on purpose
# tau of -0.22 which is about -0.4 for Pearson's r
# decrease of 0.036 ug/L of Chromium per year
```

## Nonparametric trend with covariate

* compute smooth of censored y vs x where x is not time using a Generalised Additive Model (GAM)
* GAMS are smoothed curves weighted combinations of multiple functions
* compute an ATS on residuals. Kendall's tau a test of change in residuals over time
* uses decimal time


```{r}
centrend(Dairy_Creek_Chromium$`Total Recoverable Chromium`, Dairy_Creek_Chromium$CrND, Dairy_Creek_Chromium$mean_daily_flow_cfs, Dairy_Creek_Chromium$dectime)
# has plot of data with ND and GAM smooth
# ATS on residuals shwos trend (p-value of 0.03) with Kendall's tau showing a decrease of 0.0225 per year
# second plot shows resids from GAM
```

## Seasonal Kendall test on censored data

* computes ATS line and test for each season separately
* combines in to one overall seasonal Kendall test
* test of consistent trend (if one season shows increasing significant trend then second shows decreasing, then cancel out and no overall SK trend)
* does not compare data across different seasons
* test statistic for each season is the Mann-Kendall test. sum up for overall

```{r}
# time first
censeaken(Dairy_Creek_Chromium$dectime,
          Dairy_Creek_Chromium$`Total Recoverable Chromium`,
          Dairy_Creek_Chromium$CrND,
          group = Dairy_Creek_Chromium$Season,
          seaplots = TRUE) # for seasonal graphs

# significant in dry season
# not signifiant in wet
# significant overall slope is -0.036 per year
# histogram shows S permutations - time variable randomly shuffled thousands of times and DV is reassigned, then S computed for each shuffle
```


