---
title: "Section 13. Multivariate Methods with Nondetects"
output: html_notebook:
  toc: true
---

```{r}
# libraries-----------------------------------------------------------
source("NADA2_online_3_9/Class Data/Loadlibs.R")
library(tidyverse)
library(glue)
library(bestglm)
library(rms)
library(readxl)
```

```{r}
# data --------------------------------------------------------
data_folder <- "NADA2_online_3_9/Class Data"

FishDDT <- read_excel(glue("{data_folder}/FishDDT.xls"))
load(glue("{data_folder}/FishDDTalt.rdata"))

# exercises
load(glue("{data_folder}/Markers.rdata"))
```

# Binary multivariate methods for censored data

Three methods, all three put them on a scale that equalises the DV so that each has the same impact
* convert data to binary detect/nondetect, 0 or 1
* re-censor all values below the highest DL then rank within each variable and perform multivariate analyses on the ranks (ordinal approach)
* use u-scores to develop relationships within multiply-censored variables

## Types of multivariate models

Inter-dependence (all variables equivalent) -- correlations between all the variables
* NMDS - non metric multidimensional scaling
* Clustering

Dependence (response and explanatory variables)
* ANOSIM - like a multivariate version of Kruskal-Wallis
* Seriation (nonparametric correlation / trend test) - multivariate version of Kendall's tau

## Distance or similarity matrix

* big set of correlation coefficients
* quantifies the similarity or distance between rows or between columns
* matrix has triangular shape
* distance measures include Euclidean and Bray-Curtis distances between points in multivariate space

```{r}
# measures for censored data
tribble(~distance_measure, ~characteristic,
        "simple matching (binomial)", "above/below RL",
        "Euclidean distance (ordinal)", "ranks",
        "Euclidean distance (Wilcoxon)", "u-scores")
```

## Dataset

* orthogonal and parallel versions of DDD, DDE and DDT
* `Age` (young v mature) and `Date` (decimal time)
* indicators where 1 is a nondetect

```{r}
summary(FishDDT)
```

## NMDS on simple matching distances

```{r}
DDTcen <- FishDDT %>% 
  dplyr::select(oD_LT1:pT_LT1)

binaryMDS(DDTcen, FishDDT$Age, title = "NMDS of DDT in fish")

# differentiation between young and mature fish along x axis
# overlay due to ties between pairs
# distances on triangular distance based on ranks
```

## ANOSIM - test group differences in patterns of 0/1s

```{r}
DDTdissim <- binaryDiss(DDTcen) # dissimilatity matrix without grouping variable

# from vegan package
ano_ddt <- anosim(DDTdissim, FishDDT$Age) # dissimilatity matrix as one of the inputs
ano_ddt

# anosim permutation test compares test statistic to lots of permutations after randomisation

anosimPlot(ano_ddt) # hist of permuted values - the test statistic is higher than all permutations
# pvalue is 0.001 as highest in 1000 permutations, so reject H0 that groups are the same and can say they are diff
```

## Clustering 0/1 data

```{r}
binaryClust(DDTcen, ncluster = 3) # young fish in left most, the others are mature
```

## Trend analysis for binary data

```{r}
time_dist <- dist(FishDDT$Date, method = "manhattan") # use Manhattan distance rather than Euclidean dist

# mantel test using vegan package
ddt_mannkendall <- mantel(time_dist, DDTdissim, method = "kendall", permutations = 999)
ddt_mannkendall 
# Mantel statistic based on Kendall's rank correlation tau = 0.151
# shows quantiles of permutations

# significant in the 0/1 pattern over time - like ANOSIM higher than almost all the permutations
densityplot(permustats(ddt_mannkendall), main = "Multivariate Mann-Kendall Test")
```

# Ordinal multivariate methods

* compute the distance measure then compute the same multivariate procedures as the binary approach
* more precision than the binary approach
* Euclidean distance used for distance on ranks

```{r}
new_fish <- FishDDT %>% 
  dplyr::select(-Age, -Date)

ranks_ddt <- ordranks(new_fish, paired = FALSE) # will censor at highest DL in column
euclid_ddt <- dist(ranks_ddt) # Euclidean is default
euclid_ddt # for example sites 1, 5 and 6 have young fish and are similar with low numbers
```

## NMNDS on Euclidean rank matrix
```{r}
# vegan package
ddt_eumds <- metaMDS(euclid_ddt)
ordiplot(ddt_eumds, type = "text", main = "NMDS of rank of DDT concentrations") # there is an error but the plot still gets produced

# points not on top of eachother
```

## ANOSIM test on rakned group differences

```{r}
rnk_ano <- anosim(euclid_ddt, FishDDT$Age)
rnk_ano #  significant but slightly higher

anosimPlot(rnk_ano)
```

## Cluster on ranks of concentrations
```{r}
rankclust <- hclust(euclid_ddt, method = "average")
plot(rankclust)
# young fish are the rows 1,6, 24, 25, 5, 26
```

## Trend test on ranks

* Are the distances between ranks correlated with time? 

```{r}
time_dist # from earlier - manhattan
ddt_ranks <- mantel(time_dist, euclid_ddt, method = "kendall", permutations = 999)
ddt_ranks # higher than all the other results, significant

densityplot(permustats(ddt_ranks), main = "Multivariate Mann-Kendall Test")
```

## Multivariate methos on u-scores (for data with multiple reporting limits)

* compute the distance measure
* compute the same multivariate procedures

```{r}
# slightly altered to add new detection limits - low and high end of intervals
FishDDTalt
```

# u-scores

* a measure of order within a variable
* the sum of the number of observations that the ith value is above minus the number ith is below
* the basis of the Mann-Whitney, Kendall's tau and other methods
* Euclidean distance on either the u-scores or the ranks of the u-scores
* author recommends ranks rather than scores

Functions
* `uscores()` computes with 0/1 indicator
* `uscoresi()` computes using low, high format
* `uMDS()` computes Euclidean distances between scores (or ranked scores) then plots


## Computing u-scores for NMDS

```{r}
alt <- FishDDTalt %>% 
  dplyr::select(-Age, -Date, -Site)

u_scores <- uscoresi(alt, paired = FALSE) # ranks of the uscores
u_euclid_ddt <- dist(u_scores) # Euclidean_dist
uMDS(u_scores, group = FishDDTalt$Age, title = "NMDS of rank(u-scores) for DDT concentrations") 

u_sc2 <- uscoresi(alt, paired = FALSE, rnk = FALSE) # uscores not ranks 
uMDS(u_sc2, group = FishDDTalt$Age, title = "NMDS of u-scores (not ranks) for DDT concentrations")
```

## ANOSIM test on uscore group differences

```{r}
u_ano <- anosim(u_euclid_ddt, FishDDT$Age)
u_ano # lower p-value than with ranks at highest DL

anosimPlot(u_ano)
```

## Cluster analysis using uscores

```{r}
uclust <- hclust(u_euclid_ddt, method = "average")
plot(uclust)
```

## Trend analysis using uscores

* triangles are dist between uscores of concentrations and dist btw times (manhattan btw years)

```{r}
ddt_utrend <- mantel(time_dist, u_euclid_ddt, method = "kendall", permutations = 999)
ddt_utrend
# very significant
# trend in pattern of DDT and its degradation of prducts over time
# test statistic of 0.432 is over four times the 99th percentile of permutations
densityplot(permustats(ddt_utrend), main = "Multivariate Mann-Kendall Test")
```

# Exercises

## Dataset

* Microbial source tracking (MST) markers to detect fecal pollution
* Interval censored format where (0 to MDL) indicates values below DL
* non-zero lower ends indicate either MDL to QL or above QL
* Entero1A lo and hi general fecal pollution indicators

To do:
* test whether pattern of six MST markers plus Entero1A indicators differ amongst five sites using ANOSIM
* test whether there is a trend between the MST markers v Entero1A using mantel test

```{r}
Markers # note that one row has NAs

# new dataset without site
m2_new <- Markers %>% 
  drop_na()

m2 <- m2_new %>% 
  dplyr::select(-Site_Name) 

sites <- m2_new$Site_Name

m2_uscores <- uscoresi(m2, paired = FALSE) # ranks of the uscores
m2_euclid <- dist(m2_uscores) # Euclidean_dist
```

```{r}
m2_ano <- anosim(m2_euclid, sites)
m2_ano # lower p-value than with ranks at highest DL

anosimPlot(m2_ano) # significant and higher than most of the permutations

uMDS(m2_uscores, group = sites, title = "NMDS of rank(u-scores) for markers and entero") 
# Miami Central mostly on the right
# Bakers Haulover mostly on left
```

Test trend

```{r}
# mst data
m3 <- m2_new %>% 
  select(1:12) # remove sites and entero

m3_uscores <- uscoresi(m3)
m3_euclid <- dist(m3_uscores) # euclidean

# entero data
ent <- m2_new %>% 
  select(starts_with("Entero"))

ent_uscores <- uscoresi(ent)
ent_euclid <- dist(ent_uscores) # euclidean

# mantel analysis
m3_utrend <- mantel(m3_euclid, ent_euclid, method = "kendall", permutations = 999)
m3_utrend # significant
densityplot(permustats(m3_utrend), main = "Multivariate Mann-Kendall Test")

# plot
tibble(ent_euclid,
       m3_euclid,
       sites = rep(sites, 14)) %>% 
  ggplot(aes(ent_euclid, m3_euclid)) + 
  geom_point(aes(colour = sites)) +
  geom_smooth() + 
  theme(legend.position = "bottom") +
  labs(title = "Correlation of distance matrix of rank(uscores)",
       x = "Entero Euclidean distances",
       y = "MST Euclidean distances",
       colour = NULL)


```

# Extra

Determine which MST markers have the highest correlation with Entero1A values

```{r}
bioenv(comm = ent_euclid, env = m3_uscores, method = "kendall")
# the highest concentration is the HF183 marker

tibble(ent = ent_uscores,
       hf183 = m3_uscores[,4]) %>% # the fourth column
         ggplot(aes(ent, hf183)) +
         geom_point() 
```

