---
title: "Section 8. Two-Group Tests with Nondetects"
output: html_notebook:
  toc: true
---

```{r}
# libraries-----------------------------------------------------------
source("NADA2_online_3_9/Class Data/Loadlibs.R")
library(tidyverse)
library(glue)
```

```{r}
# data --------------------------------------------------------
data_folder <- "NADA2_online_3_9/Class Data"

zinc <- read.csv(glue("{data_folder}/Zinc.txt"), sep = "") %>% 
  as_tibble()

load(glue("{data_folder}/TCE2.RData"))
```

# Methods for testing differences between groups

* Parametric tests - tests for means - must assume distributions that matches data 
* Non-parametric - tests for percentiles - no shape necessary
* Simpler tests - convert to above or below detection limit - non param

# Parametric ('t-test' by regression)

* Regression of concentrations v group if
* Concentration = intercept + slope * factor
* is one group different to the other? HO of not diff
* parametric and assumes normal dist 
* could take logs 

```{r}
censummary(zinc$Zn, zinc$ZnLT, factor(zinc$Zone))
```



```{r}
cboxplot(zinc$Zn, zinc$ZnLT, zinc$Zone, minmax = TRUE, show = TRUE)
```

## With MLE in normal units

```{r}
cen2means(zinc$Zn, zinc$ZnLT, zinc$Zone, LOG = FALSE) 
# does not find evidence that the two means differ
# but not a normal distribution -- estimates < 0
```
## With a permutation test 

```{r}
cenperm2(zinc$Zn, zinc$ZnLT, zinc$Zone) # better and values make sence
# still reject H0 and can say that they are diff
```

## With MLE in log units

```{r}
cen2means(zinc$Zn, zinc$ZnLT, zinc$Zone, LOG = TRUE) 
# still rejects log normality
# finds no difference in means again
```

# Non parametric tests for censored data

* extend the Wilcoxon rank-sum tests to censored data
* score and rank are essentially Kaplan-Meier
* R version is called the 'Peto-Peto' test
* test of whether the survival curves (KM percentages) are the same for all groups

```{r}
cen1way(zinc$Zn, zinc$ZnLT, zinc$Zone)
# low p-value - can reject H0 and say that they are diff
```

```{r}
# the emperical cdf
cen_ecdf(zinc$Zn, zinc$ZnLT, zinc$Zone) # top part of the box (above 35th percentile) is diff
```

# Simpler methods

* Might use if you want to give a simple demonstration, but otherwise don't do
* Contingency tables for less than highest DL and higher than DL - loses information but simple

```{r}
# chi squared value finds a difference in the data
chisq.test(xtabs(~TCE2$Density + TCE2$Below5Cens)) 
```


