---
title: "Section 6. Intervals (Confidence, Prediction, Tolerance)"
output: html_notebook:
  toc: true
---

```{r}
# libraries-----------------------------------------------------------
source("NADA2_online_3_9/Class Data/Loadlibs.R")
library(tidyverse)
library(glue)
```

```{r}
# data --------------------------------------------------------
data_folder <- "NADA2_online_3_9/Class Data"

data("ShePyrene")
```


# Confidence intervals

## Kaplan-Meier

* t.LCL and UCL assume normality
* Pct.LCL and UCL (percentile)
* BCa.LCL and UCL (biased correction) -- do not use if more than 40% non-detect

Bootstrap CI reflects the skewness properly.

Kaplan-Meier uses the data, while MLE uses the model.

```{r}
# envstats
km_ci <- enparCensored(ShePyrene$Pyrene,
              ShePyrene$PyreneCen,
              ci = TRUE, # two sided by default
              ci.method = "bootstrap",
              n.bootstraps = 5000)

print(km_ci)
```

## MLE

Would run if more sure that it is lognormally distributed.

```{r}
# nada -- assumes lognormal by default
pyr_mle_ci <- cenmle(ShePyrene$Pyrene,
                     ShePyrene$PyreneCen,
                     conf.int = 0.95)

mean(pyr_mle_ci)
print(pyr_mle_ci)
```

```{r}
# assuming gaussian
cenmle(ShePyrene$Pyrene,
                     ShePyrene$PyreneCen,
                     dist = "gaussian")

```

```{r}
# envstats
pyr_mle_envstats_ci <- elnormAltCensored(ShePyrene$Pyrene,
                                      ShePyrene$PyreneCen,
                                      ci = TRUE,
                                      ci.method = "bootstrap",
                                      n.bootstraps = 5000)

print(pyr_mle_envstats_ci)
```

## ROS

```{r}
# envstats
pyr_ros_envstats_ci <- elnormAltCensored(ShePyrene$Pyrene,
                                      ShePyrene$PyreneCen,
                                      method = "rROS",
                                      ci = TRUE,
                                      ci.method = "bootstrap",
                                      n.bootstraps = 5000)

print(pyr_ros_envstats_ci)
```

# Prediction intervals

Want to see if new observations have changed from background (what is already there). Where you would be 95% sure of where new observations would fall.

## MLE

* Note that the normal LPL here is negative, so you wouldn't use it
* Would choose which dist using BIC - see section 5

```{r}
# NADA
cenPredInt(ShePyrene$Pyrene, ShePyrene$PyreneCen) # as a df!!!!
```

## ROS

The more observations that you want to predict for, the wider the observation

```{r}
cenPredInt(ShePyrene$Pyrene, ShePyrene$PyreneCen, newobs = 2, method = "rROS")
```


# Tolerance intervals

* 90th percentile of data
* 95% UTL so that we're 95% confident that the 90th percentile should not be higher than this number
* the plot helps to choose which value to choose
* smallest BIC is best - here it is the lognormal

## Lognormal, gamma and normal

```{r}
# nada
cenTolInt(ShePyrene$Pyrene, ShePyrene$PyreneCen)
```


## Lognormal MLE

Takes logs, computes stats, then transforms back

```{r}
# envstats
lnorm_mle_envstats_ti <- eqlnormCensored(ShePyrene$Pyrene, ShePyrene$PyreneCen, p = 0.9, ci = TRUE, ci.type = "upper")
print(lnorm_mle_envstats_ti)
```

## Gamma MLE
```{r}
dat_gamma <- ShePyrene$Pyrene^(1/3) # cube root
obj_gamma <- eqnormCensored(dat_gamma, ShePyrene$PyreneCen, p = 0.9, ci = TRUE, ci.type = "upper")
pct_gamma <- obj_gamma$quantiles^3 # cube the 90th percentile
ti_gamma <- (obj_gamma$interval$limits[2])^3 # cube the upper limit

pct_gamma
ti_gamma
```

