---
title: "Section 12. Logistic Regression"
output: html_notebook:
  toc: true
---

```{r}
# libraries-----------------------------------------------------------
source("NADA2_online_3_9/Class Data/Loadlibs.R")
library(tidyverse)
library(glue)
library(bestglm)
library(rms)
```

```{r}
# data --------------------------------------------------------
data_folder <- "NADA2_online_3_9/Class Data"

load(glue("{data_folder}/TCELogReg.rda"))

# exercises
load(glue("{data_folder}/ReconLogistic.RData"))
```

# Binary logistic regression

* predicts the probability of getting a one (concentrations that are higher than the highest DL)
* may want to know the probability of being above the threshold for some criteria
* y is a logit (the log of the odds), so `y = log(prob(above)/prob(below))`
* only assumes that the relationship between the logit and the x variable is linear

## The dataset

* `GT5` for >= 5. the dependent variable

```{r}
TCELogReg
```

# Logistic regression

```{r}
glm1 <- glm(GT5 ~ DEPTH + PctIND + POPDEN, data = TCELogReg, family = binomial(logit))
summary(glm1)

# null devialce 182.7 -- intercept only
# resid deviance = 169.8 with 243 df -- this model
# AIC 177.8

```

```{r}
# the null model
glm0 <- glm(GT5 ~ 1, data = TCELogReg, family = binomial(logit))
summary(glm0)

# compare null model to the original model using an overall Likelihood ratio test G
anova(glm0, glm1, test = "Chisq") # p-value shows that the second model is better
```

## Model selection

```{r}
vif(glm1) # fine
```

## Test whether to transform x variables

```{r}
# from car pacakge
residualPlots(glm1, type = "deviance") # some indication that POPDEN could be transformed from pvalue
# plots themselves not great because of nature of logistic regression

```

* Box and Tidwell procedure - construct a new variable and add to regression to test
* only for continuous variables only
* if slope is significant, then transform variable and remove originals

```{r}
bt_tce <- TCELogReg %>% 
  # only DEPTH and POPDEN - don't transform  percentage variables
  mutate(bt_depth = DEPTH * log(DEPTH),
         bt_popden = POPDEN * log(POPDEN),
         sqrt_popden = sqrt(POPDEN),
         ln_popden = log(POPDEN)) 

glm2 <- glm(GT5 ~ DEPTH + PctIND + POPDEN + bt_depth + bt_popden, data = bt_tce, family = binomial(logit))
summary(glm2) # no for depth, transformed popden is significant so use that instead of POPDEN

# guide for possible transform
as.vector(1 + (glm2$coefficients["bt_popden"]/glm2$coefficients["POPDEN"])) # ratio of 0.7 
# so maybe use sqrt or log of popden?

```

```{r}
# sqrt popden
glm3 <- glm(GT5 ~ DEPTH + PctIND + sqrt_popden, data = bt_tce, family = binomial(logit))
summary(glm3) # AIC 175

```

```{r}
glm4 <- glm(GT5 ~ DEPTH + PctIND + ln_popden, data = bt_tce, family = binomial(logit))
summary(glm4) # AIC 173 better
```

# Which variables to keep in the model

* AIC = 2*num_parameters + residual deviance


```{r}
# removing PctIND
glm5 <- glm(GT5 ~ DEPTH + ln_popden, data = bt_tce, family = binomial(logit))
summary(glm5) # AIC 171 
```


```{r}
anova(glm4, glm5, test = "Chisq") # p-value shows that the second model is better
# residual deviance in second is lower and one fewer df. 
# prob of getting that deviance if H0 (of simpler model being better) is 0.86 so no evidence to reject H0
```

```{r}
# removing DEPTH
glm6 <- glm(GT5 ~ ln_popden, data = bt_tce, family = binomial(logit))
summary(glm6) # AIC 170

```

```{r}
anova(glm4, glm6, test = "Chisq") # nothing much lost by dropping EPTH and PctInd
anova(glm5, glm6, test = "Chisq") # nothing lost when dropping DEPTH

```

## Automated model selection

```{r}
TCE_best <- bt_tce %>% 
  dplyr::select(PctIND, DEPTH, ln_popden, GT5) # requires dataset where the last column is the binary variable

bestglm(TCE_best, family = binomial(logit), IC = "AIC")
```

## Interpreting coefficients

* slope is 1.25 and variable is logged
* as population density increases, the probability of getting a 1 (exceeding) increases
* for a unit increase in log(POPDEN), the log-odds increase by 1.25, which is a 3.49 multiplier

```{r}
# best model
summary(glm6)

exp(glm6$coefficients["ln_popden"]) # e^1.2515 = 3.495
```

* supposing `glm5` was chosen
* the slope of -0.0013 for DEPTH says the probability of TCE ?= 5 decreases as depth increases
* for each unit increase in depth, there is an (e^-0.0013) - 0.9987 multiplier (or a 0.13% decrease) in the odds (p/(1-p)) for TCE >= 5.

```{r}
summary(glm5)

exp(glm5$coefficients["DEPTH"]) # e^1.158548 = 0.9987
```

## Evaluate the chosen model

* R^2 might help to decide whether to use the log of a variable or not (or sqrt or not)
* Nagelkerke r^2 is a rescaled likelihood in logistic regression and tends to be a bit smaller than R^2 in standard regression

### R^2

```{r}
# from rms package
lrm6 <- lrm(GT5 ~ ln_popden, data = TCE_best)
lrm6
```

### Plots and measures

```{r}
# plots from rms
d6 <- datadist(TCE_best$ln_popden, TCE_best$GT5)
options(datadist = "d6")
## not sure if this works
# plot(Predict(lrm6))


d5 <- datadist(TCE_best$ln_popden, TCE_best$DEPTH, TCE_best$GT5)
options(datadist = "d5")
lrm5 <- lrm(GT5 ~ ln_popden + DEPTH, data = TCE_best)
# # also doesn't work
# plot(Predict(lrm5)) 

```

```{r}
# diff graph
pof1 <- exp(glm6$linear.predictors) / (1-exp(glm6$linear.predictors))

psort <- order(TCE_best$ln_popden) # order
plot(TCE_best$ln_popden,
     TCE_best$GT5,
     xlab = "ln(Population Density)",
     ylab = "Prob TCE >= 5")
lines(TCE_best$ln_popden[psort],
      pof1[psort],
      col = "blue")
```

### Predictive ability

Best are:
* best are Kendall's tau-b (tau adjusted for ties) - higher is better - use cor.test()
* AUC - area under the curve - higher is better
* Brier score - mean squared error of prediction - lower is better


* `lrm()` Brier: Brier score
* `lrm()` C: AUC
* `lrm()` Dxy = Somer's Dxy
* `lrm()` tau-a =  tau-a
* `cor.test()` Kendall's tau: tab-b

```{r}
lrm6 # from earlier

cor.test(glm6$linear.predictors, TCE_best$GT5, method = "kendall")
```

# Exercise

Atrazine measured in streams. Reporting limit of 1 (`GT_1`). Land annd climate variables:

* `APPLIC` - amount of pesticide applied
* `CORNpct` - pct of basin planted in corn
* `SOILGP` - soil permeability
* `PRECIP` - amount of recent precipitation
* `DYPLANT` - days since planting
* `FPCTL` - percentile of streamflow
* `TEMP` - supposely not meant to use this, but temp, not sure if C or F.

Build the best logistic regression model.

```{r}
ReconLogistic
```

```{r}
rl1 <- glm(GT_1 ~ . - TEMP, data = ReconLogistic, family = binomial(logit))
summary(rl1) # AIC 259.9

# soil, precip and applic prob not great

vif(rl1) # fine
```

```{r}
# compare to intercept only
rl0 <- glm(GT_1 ~ 1 - TEMP, data = ReconLogistic, family = binomial(logit))
summary(rl0) # AIC 560

anova(glm0, glm1, test = "Chisq") # intercept only no good
```


```{r}
# see whether to maybe transform
residualPlots(rl1, type = "deviance") # perhaps DYPLANT and SOILGP
```

```{r}
# box tidwell
bt_rl <- ReconLogistic %>% 
  # only DEPTH and POPDEN - don't transform  percentage variables
  mutate(bt_dyplant = DYPLANT * log(DYPLANT),
         bt_soil = SOILGP * log(SOILGP),
         bt_precip = PRECIP * log(PRECIP)) 

rl2 <- glm(GT_1 ~ ., data = bt_rl, family = binomial(logit))
summary(rl2) # yes for dyplant, yes for soil # AIC 236

# guide for possible transform
as.vector(1 + (rl2$coefficients["bt_dyplant"]/rl2$coefficients["DYPLANT"])) #0.85
as.vector(1 + (rl2$coefficients["bt_soil"]/rl2$coefficients["SOILGP"])) # 0.46
as.vector(1 + (rl2$coefficients["bt_precip"]/rl2$coefficients["PRECIP"])) # 81.49

vif(rl2) # bad vif -- just checking wouldn't use anyway
```

```{r}
# use bt versions
rl3 <- glm(GT_1 ~ CORNpct + bt_dyplant + FPCTL + APPLIC + bt_dyplant + bt_soil, data = bt_rl, family = binomial(logit))
summary(rl3) # worse!!! AIC = 267
```

```{r}
# drop APPLIC
rl4 <- glm(GT_1 ~ CORNpct + PRECIP + FPCTL + DYPLANT + SOILGP, data = ReconLogistic, family = binomial(logit))
summary(rl4) # AIC 258.7
```

```{r}
# drop SOILGP
rl5 <- glm(GT_1 ~ CORNpct + PRECIP + FPCTL + DYPLANT, data = ReconLogistic, family = binomial(logit))
summary(rl5) # AIC 257.7 and everything signif
```

```{r}
# drop CORNpct just to check
rl6 <- glm(GT_1 ~ PRECIP + FPCTL + DYPLANT, data = ReconLogistic, family = binomial(logit))
summary(rl6) # AIC 260.6 worse

anova(glm6, glm5, test = "Chisq") # worse
```

```{r}
# automated selection
bestglm(ReconLogistic %>% 
          dplyr::select(-TEMP), # not meant to use TEMP
        family = binomial(logit),
        IC = "AIC") # the same model

```




```{r}
# best
summary(rl5)

# other scores
lrm(GT_1 ~ CORNpct + PRECIP + DYPLANT + FPCTL, data = ReconLogistic)
# Brier 0.81, AUC 0.932

cor.test(rl5$linear.predictors, ReconLogistic$GT_1, method = "kendall")
# tau-b 0.591
```

